---
title: "Simulation from a Pooled Logistic Regression Model"
subtitle: "Discrete Survival Analysis & Sequential Propensity Scores"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    highlight-style: github
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    highlight-style: github
    colorlinks: true
    geometry:
      - top=25mm
      - bottom=25mm
      - left=25mm
      - right=25mm
execute:
  warning: false
  message: false
  echo: true
---

## Background

### Discrete Survival Analysis

Suppose that event times can only happen at times $\tau_1, \dots, \tau_K, \infty$,
where $\infty$ means that no event was observed within $[0, \tau_K]$.
The **survival function** at time $k$ is defined analogously to the continuous case:

$$
S(k) = \Pr(T > k).
$$

The **discrete-time hazard function** at time $k$ is a conditional probability:

$$
h(k) = \Pr(T = k \mid T \geq k).
$$

Some useful relationships follow from these definitions:

$$
\begin{aligned}
\Pr(T = k) &= h(k)\, S(k-1), \\
S(k)       &= S(k-1)\bigl(1 - h(k)\bigr), \\
S(k)       &= \prod_{j=1}^{k} \bigl(1 - h(j)\bigr).
\end{aligned}
$$

Taking logarithms yields the **cumulative hazard**:

$$
H(k) = -\log S(k) = -\sum_{j=1}^{k} \log\bigl(1 - h(j)\bigr).
$$

### Pooled Logistic Model

The discrete-time hazard is linked to covariates via a logistic model:

$$
\operatorname{logit}\, h(k \mid \mathbf{X}_{ik})
  = \operatorname{logit}\, \Pr(T = k \mid T \geq k,\, \mathbf{X}_{ik})
  = \alpha_k + \boldsymbol{\beta}^\top \mathbf{X}_{ik},
$$

where $\alpha_k$ are **time-specific intercepts** (encoding the baseline hazard)
and $\mathbf{X}_{ik}$ are (possibly time-varying) covariates.

### Sequential Propensity Score

Let $T$ be the time of treatment initiation, $T \in \{1, \dots, K, \infty\}$,
and define $A_k = \mathbf{1}[T = k]$. The **sequential propensity score** is

$$
e_k(\bar{L}_k) = \Pr(A_k = 1 \mid \bar{A}_{k-1} = 0,\, \bar{L}_k).
$$

Because $\{T \geq k\} \equiv \{\bar{A}_{k-1} = 0\}$, we have the equivalence

$$
\boxed{e_k(\bar{L}_k) = h(k \mid \bar{L}_k).}
$$

Estimating a sequential propensity score is therefore equivalent to estimating
the discrete-time hazard of treatment. The joint probability of a treatment
history up to month $k$ factorises as

$$
\Pr(T = k \mid \bar{L}_k)
  = h(k \mid \bar{L}_k) \prod_{j < k} \bigl(1 - h(j \mid \bar{L}_j)\bigr),
$$

which is the foundation for **inverse probability weights** in longitudinal
causal inference.

---

## Setup

```{r setup}
library(tidyverse)

set.seed(42)
```

---

## Step 1 — Define Parameters

We choose $K = 10$ discrete time intervals and specify:

- **Baseline hazard** $\alpha_k$: a linearly increasing sequence on the logit
  scale (risk increases over time).
- **Covariate effects**: $\beta_1 = 0.8$ for a continuous covariate $X_1$ and
  $\beta_2 = -0.5$ for a binary covariate $X_2$.

```{r params}
n      <- 2000                              # number of individuals
K      <- 10                                # number of discrete time points

# Time-specific intercepts alpha_k (baseline log-odds of the hazard)
alpha  <- seq(-3, -1.5, length.out = K)

# Covariate coefficients
beta_x1 <-  0.8    # continuous covariate
beta_x2 <- -0.5    # binary covariate
```

---

## Step 2 — Generate Baseline Covariates

Each individual has two fixed baseline covariates:

$$
X_1 \sim \mathcal{N}(0,\,1), \qquad X_2 \sim \text{Bernoulli}(0.4).
$$

```{r covariates}
X1 <- rnorm(n, mean = 0, sd = 1)      # continuous
X2 <- rbinom(n, size = 1, prob = 0.4) # binary
```

---

## Step 3 — Simulate Discrete Event Times

At each interval $k$, the discrete-time hazard for individual $i$ is

$$
h(k \mid X_i) = \operatorname{expit}(\alpha_k + \beta_1 X_{1i} + \beta_2 X_{2i}).
$$

We draw $A_k \sim \text{Bernoulli}\bigl(h(k \mid X_i)\bigr)$ **only for those
still at risk** ($T \geq k$). The event time is the first $k$ at which $A_k = 1$.
Individuals for whom no event occurs by $\tau_K$ are assigned $T = \infty$.

```{r simulate}
simulate_event_times <- function(n, K, alpha, beta_x1, beta_x2, X1, X2) {

  T_event <- rep(Inf, n)      # Inf = no event within [tau_1, tau_K]
  at_risk <- rep(TRUE, n)     # everyone starts at risk

  for (k in seq_len(K)) {

    # Linear predictor for interval k
    eta_k <- alpha[k] + beta_x1 * X1 + beta_x2 * X2

    # Discrete-time hazard h(k | X_i)
    h_k   <- plogis(eta_k)

    # Bernoulli trial only for individuals still at risk
    A_k   <- rbinom(n, size = 1, prob = h_k * at_risk)

    # Record event time for newly treated individuals
    T_event[A_k == 1 & at_risk] <- k

    # Remove newly treated from risk set
    at_risk <- at_risk & (A_k == 0)
  }

  return(T_event)
}

T_event <- simulate_event_times(n, K, alpha, beta_x1, beta_x2, X1, X2)

cat("Event rate (within follow-up):", round(mean(T_event <= K), 3), "\n")
```

---

## Step 4 — Introduce Independent Censoring

We draw a random censoring time $C \sim \text{Uniform}\{1, \dots, K\}$,
independently of covariates and the event process. The observed data are

$$
\tilde{T}_i = \min(T_i, C_i), \qquad
\delta_i    = \mathbf{1}[T_i \leq C_i].
$$

```{r censoring}
C_time <- sample(1:K, n, replace = TRUE)   # censoring times

T_obs  <- pmin(T_event, C_time)            # observed time
delta  <- as.integer(T_event <= C_time)    # event indicator

cat("Observed event rate:", round(mean(delta), 3), "\n")
```

### Visualisation of Simulated Event Times

The plots below summarise the distribution of simulated times across the $K$
intervals. The left panel shows **true event times** $T_i$ (those with
$T_i \leq K$ are events; $T_i = \infty$ is shown as a separate "No event"
bar). The right panel breaks the **observed times** $\tilde{T}_i$ down by
whether the individual experienced the event ($\delta = 1$) or was censored
($\delta = 0$), illustrating how censoring redistributes observations across
intervals.

```{r plot-simulated-times, fig.width=11, fig.height=4.5}
library(patchwork)

# --- Panel A: distribution of true event times ---
true_times_df <- tibble(
  T_true = T_event,
  label  = if_else(T_event <= K, paste0("k = ", T_event), "No event")
) |>
  mutate(label = factor(label,
                        levels = c(paste0("k = ", 1:K), "No event"))) |>
  count(label) |>
  mutate(pct = n / sum(n) * 100)

p1 <- ggplot(true_times_df, aes(x = label, y = n,
                                fill = label == "No event")) +
  geom_col(colour = "white", width = 0.75) +
  geom_text(aes(label = sprintf("%.1f%%", pct)),
            vjust = -0.4, size = 3) +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "grey60"),
                    guide  = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "A — True event times",
    x     = "Interval / Status",
    y     = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

# --- Panel B: observed times stratified by event / censoring ---
obs_times_df <- tibble(
  interval = T_obs,
  status   = factor(delta, levels = c(1, 0),
                    labels = c("Event (\u03b4 = 1)", "Censored (\u03b4 = 0)"))
) |>
  count(interval, status)

p2 <- ggplot(obs_times_df, aes(x = factor(interval), y = n, fill = status)) +
  geom_col(colour = "white", width = 0.75, position = "stack") +
  scale_fill_manual(values = c("Event (\u03b4 = 1)"    = "steelblue",
                               "Censored (\u03b4 = 0)" = "salmon")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(
    title = "B — Observed times by event/censoring status",
    x     = "Observed interval k",
    y     = "Count",
    fill  = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

p1 + p2
```

---

## Step 5 — Expand to Person-Period Format

The key structural step: each individual $i$ contributes **one row per interval
they remained at risk**, up to and including $\tilde{T}_i$.
The binary outcome is

$$
Y_{ik} =
\begin{cases}
1 & \text{if } k = \tilde{T}_i \text{ and } \delta_i = 1, \\
0 & \text{otherwise.}
\end{cases}
$$

```{r long-format}
expand_to_long <- function(id, T_obs, delta, X1, X2) {

  map_dfr(seq_along(id), function(i) {

    n_periods <- T_obs[i]
    if (n_periods == 0) return(NULL)

    intervals <- seq_len(n_periods)

    tibble(
      id       = id[i],
      interval = intervals,
      Y        = as.integer(intervals == T_obs[i] & delta[i] == 1),
      X1       = X1[i],
      X2       = X2[i]
    )
  })
}

long_df <- expand_to_long(
  id    = 1:n,
  T_obs = T_obs,
  delta = delta,
  X1    = X1,
  X2    = X2
)

cat("Rows in long format:", nrow(long_df),
    "(expected ~", sum(T_obs), ")\n")

glimpse(long_df)
```

---

## Step 6 — Fit the Pooled Logistic Regression

We model `interval` as a **factor** so that each level gets its own intercept
$\hat{\alpha}_k$, recovering the full shape of the baseline hazard without
imposing any smoothness assumption.

$$
\operatorname{logit}\, h(k \mid X_i) = \alpha_k + \beta_1 X_{1i} + \beta_2 X_{2i}.
$$

```{r fit-plr}
long_df <- long_df |>
  mutate(interval_f = factor(interval, levels = 1:K))

fit_plr <- glm(
  Y ~ interval_f + X1 + X2 - 1,   # -1: factor supplies all intercepts
  data   = long_df,
  family = binomial(link = "logit")
)

summary(fit_plr)
```

---

## Step 7 — Compare Estimated vs. True Parameters

```{r compare-params}
coef_est <- coef(fit_plr)
ci       <- confint.default(fit_plr)  # Wald CIs

results <- tibble(
  parameter  = names(coef_est),
  true_value = c(alpha, beta_x1, beta_x2),
  estimate   = coef_est,
  lower      = ci[, 1],
  upper      = ci[, 2]
) |>
  mutate(
    covers_truth = lower <= true_value & true_value <= upper
  )

knitr::kable(results, digits = 3,
             caption = "Estimated vs. true parameters (Wald 95% CI)")
```

All confidence intervals should cover the true values, confirming that the
pooled logistic model correctly recovers the data-generating parameters.

---

## Step 8 — Recover and Plot the Baseline Hazard

Back-transforming the $\hat{\alpha}_k$ estimates via $\operatorname{expit}$
recovers the estimated baseline hazard $\hat{h}_0(k)$.

```{r plot-hazard}
baseline <- results |>
  filter(str_detect(parameter, "interval_f")) |>
  mutate(
    k       = 1:K,
    h_true  = plogis(true_value),
    h_est   = plogis(estimate),
    h_lower = plogis(lower),
    h_upper = plogis(upper)
  )

ggplot(baseline, aes(x = k)) +
  geom_ribbon(aes(ymin = h_lower, ymax = h_upper),
              alpha = 0.2, fill = "steelblue") +
  geom_line(aes(y = h_est,  colour = "Estimated"), linewidth = 1) +
  geom_line(aes(y = h_true, colour = "True"),
            linewidth = 1, linetype = "dashed") +
  geom_point(aes(y = h_est), colour = "steelblue", size = 2) +
  scale_colour_manual(
    values = c("Estimated" = "steelblue", "True" = "firebrick")
  ) +
  scale_x_continuous(breaks = 1:K) +
  labs(
    title   = "Baseline Hazard: True vs. Estimated",
    x       = "Interval k",
    y       = expression(h[0](k) == expit(alpha[k])),
    colour  = NULL,
    caption = "Shaded band = pointwise Wald 95% CI"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

---

## Step 9 — Recover and Plot the Survival Function

The discrete-time survival function is recovered via the product-limit formula:

$$
\hat{S}(k) = \prod_{j=1}^{k} \bigl(1 - \hat{h}_0(j)\bigr).
$$

```{r plot-survival}
baseline <- baseline |>
  mutate(
    S_true = cumprod(1 - h_true),
    S_est  = cumprod(1 - h_est)
  )

ggplot(baseline, aes(x = k)) +
  geom_step(aes(y = S_est,  colour = "Estimated"), linewidth = 1) +
  geom_step(aes(y = S_true, colour = "True"),
            linewidth = 1, linetype = "dashed") +
  scale_colour_manual(
    values = c("Estimated" = "steelblue", "True" = "firebrick")
  ) +
  scale_x_continuous(breaks = 1:K) +
  labs(
    title  = "Baseline Survival Function: True vs. Estimated",
    x      = "Interval k",
    y      = "S(k)",
    colour = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

---

## Summary

The table below recaps the full simulation and estimation pipeline.

| Step | Action | Key object |
|------|--------|------------|
| 1 | Define $\alpha_k$, $\beta_1$, $\beta_2$ | `alpha`, `beta_x1`, `beta_x2` |
| 2 | Draw baseline covariates $X_1$, $X_2$ | `X1`, `X2` |
| 3 | Sequential Bernoulli trials $\Rightarrow$ event times | `T_event` |
| 4 | Apply independent censoring | `T_obs`, `delta` |
| 5 | Expand to person-period (long) format | `long_df` |
| 6 | Fit PLR via `glm(..., binomial)` | `fit_plr` |
| 7 | Compare $\hat{\theta}$ vs. $\theta^*$ | `results` |
| 8 | Recover $\hat{h}_0(k) = \operatorname{expit}(\hat{\alpha}_k)$ | `baseline` |
| 9 | Recover $\hat{S}(k) = \prod_{j \leq k}(1-\hat{h}_0(j))$ | `baseline` |
